Classical Planning with Simulators: \\ Results on the Atari Video Games

### 2015-04-30 hsdip-15-paper/paper.tex
added HSDIP formatted as IJCAI submission
hsdip camera ready

\documentclass{article}
% The file ijcai15.sty is the style file for IJCAI-15 (same as ijcai07.sty).
\usepackage{ijcai15}

@@

\author{\textbf{ID:} 320  \ \ ; \ \ \textbf{Keywords:} classical planning, search, video games, application}

% \author{Nir Lipovetzky \\
%        The University of Melbourne \\
%     Melbourne, Australia\\
%     {\normalsize\url{nir.lipovetzky@unimelb.edu.au}}
%%% footnote doesn't look nice ; this looks better
%% although can't use Nir's unimelb address that is too long :-)
%% 
%    {\small\url{nirlipo@gmail.com}}
%    \And
%    Miquel Ramirez \\
%    Australian National University \\
%    Canberra, Australia\\
%   {\small\url{miquel.ramirez@gmail.com}}
%    \And
%    Hector Geffner \\
%    ICREA \&  Universitat Pompeu Fabra \\
%    Barcelona, SPAIN \\
%    {\small\url{hector.geffner@upf.edu}}
 %%%%% \thanks{firstname.lastname}
%}

\maketitle

\begin{abstract}
The Atari 2600 games supported in  the Arcade Learning Environment~\cite{bellemare:jair2013}
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  off-the-shelf as there is no  compact PDDL-model of  the games, and action effects and goals are  not
known \emph{a priori}. Indeed, there are no explicit goals, and the planner must select actions on-line while interacting 
with a simulator that returns successor states and rewards. None of this precludes  the use of  blind lookahead algorithms 
for action selection like breadth-first search or  Dijkstra's yet such methods are not effective over large  state spaces. 
We thus turn to a different class of  planning methods  introduced recently  that  have been shown to be  effective for solving
large planning problems but which do not require  prior knowledge of state transitions and costs (rewards) or goals.
The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
where  $\IW(k)$ is  a breadth-first search in which a state is  pruned when it is not the first state in the  search  to make true 
a subset of $k$ atoms. The  empirical results over \numgames\ Atari games show that  $\IW(1)$ performs at the level of UCT, 
the state-of-the-art planning method in this domain, and suggest the potential of width-based methods for planning with
simulators when factored, compact action models are not available.
\end{abstract}

\section{Introduction}

@@

We tested \IW(1) and \BFS\ over \numgames\  of the $55$ different games 
considered in \cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
We left out  \textsc{Skiing} as the reported figures apparently use a  different reward structure.}
%% check
% \footnote{
% Entries marked as  ``--'' in the table are   still running at the time of submission. 
% Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
% Eventually all of  the  $55$ games will  be included in the evaluation.}
The two  algorithms were used to play the  games in the \emph{on-line planning
setting} supported by ALE where we will compare them with the planning algorithms 
considered by BNVB; namely, breadth-first search and \UCT. ALE supports also a \emph{learning setting} 
where the goal is to learn controllers that map  states into actions without
doing any lookahead. Algorithms across  the two settings are thus not directly comparable
as they compute different things. Learning controllers appears as a more challenging problem
and it is thus not surprising that planning algorithms like \UCT\ tend to achieve a higher score than 
learning algorithms. In addition, the learning algorithms reported by BNVB tend to use 
the state of the screen pixels, while the planning algorithms,  use the state
of the RAM memory. It is not clear however whether one input representation is
more challenging than the other; for the learning algorithms, BNVB mention
that the results tend to be better for the screen inputs.

@@

Last, in the lookahead, \IW(1) and \BFS\ select an action every $5$ frames, while \UCT\  selects 
an action in every frame.  This means that in order to explore a branch 
$300$ frames deep, UCT gets to choose $300$ actions, while \IW($1$) and \BFS\ 
get to choose $60$ actions, both however using the same $300$ frames from the
budget. For this, we  followed the setup  of \BRFS\ in BNVB 
that also selects actions every $5$ frames, matching the behavior
of the emulator that requests an action also every $5$ frames. Since the 
lookahead budget is given by a maximum number of (new) frames, and the time 
is mostly taken by calls to the emulator, this may not be the best choice  
for $\IW(1)$ and \BFS\ that may therefore not be exploiting all the 
options afforded by the budget. We'll look at this further in the future.

@@

\noindent \textbf{Acknowledgments.} We thank the people involved in
the creation and maintenance of the Arcade Learning Environment that
have made this work possible. The research by Nir Lipovetzky is partially
funded by the Australian Research Council Linkage grant LP11010015.
}

>>>>

\documentclass{article}
% The file ijcai15.sty is the style file for IJCAI-15 (same as ijcai07.sty).
\usepackage{aaai}

@@

%\author{ \textbf{Keywords:} classical planning, search, video games, application}

 \author{Nir Lipovetzky \\
        { The University of Melbourne }\\
     { Melbourne, Australia}\\
     {\footnotesize\url{nir.lipovetzky@unimelb.edu.au}}
%% footnote doesn't look nice ; this looks better
% although can't use Nir's unimelb address that is too long :-)
% 
%    {\tiny\url{nirlipo@gmail.com}}
    \And
    Miquel Ramirez \\
    { Australian National University} \\
    { Canberra, Australia}\\
   {\footnotesize\url{miquel.ramirez@anu.edu.au}}
    \And
    Hector Geffner \\
    { ICREA \& U. Pompeu Fabra }\\
    { Barcelona, Spain }\\
    {\footnotesize\url{hector.geffner@upf.edu}}
 %%%% \thanks{firstname.lastname}
}

\maketitle
\begin{abstract}
The Atari 2600 games supported in the Arcade Learning
Environment~\cite{bellemare:jair2013} all feature a known initial
(RAM) state and actions that have deterministic effects. Classical
planners, however, cannot be used off-the-shelf as there is no compact
PDDL-model of the games, and action effects and goals are not known
\emph{a priori}. Indeed, there are no explicit goals, and the planner
must select actions on-line while interacting with a simulator that
returns successor states and rewards. None of this precludes the use
of blind lookahead algorithms for action selection like breadth-first
search or Dijkstra's yet such methods are not effective over large
state spaces.  We thus turn to a different class of
planning methods introduced recently that have been shown to be
effective for solving large planning problems but which do not require
prior knowledge of state transitions, costs (rewards) or goals. The
empirical results over \numgames\ Atari games show that the simplest
such algorithm performs at the level of UCT, the state-of-the-art
planning method in this domain, and suggest the potential of
width-based methods for planning with simulators when factored,
compact action models are not available.

\end{abstract}

\noindent \emph{A version of this short paper has been accepted at IJCAI \cite{lipo:ijcai15}}
\section{Introduction}

@@

We tested \IW(1) and \BFS\ over \numgames\  of the $55$ different games 
considered in \cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
We left out  \textsc{Skiing} as the reported figures apparently use a  different reward structure.}
%% check
% \footnote{
% Entries marked as  ``--'' in the table are   still running at the time of submission. 
% Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
% Eventually all of  the  $55$ games will  be included in the evaluation.}
The two  algorithms were used to play the  games in the \emph{on-line planning
setting} supported by ALE where we will compare them with the planning algorithms 
considered by BNVB; namely, breadth-first search and \UCT. ALE supports also a \emph{learning setting} 
where the goal is to learn controllers that map  states into actions without
doing any lookahead. Algorithms across  the two settings are thus not directly comparable
as they compute different things. Learning controllers appears as a more challenging problem
and it is thus not surprising that planning algorithms like \UCT\ tend to achieve a higher score than 
learning algorithms. In addition, the learning algorithms reported by BNVB tend to use 
the state of the screen pixels, while the planning algorithms,  use the state
of the RAM memory. It is not clear however whether the use of one input representation is
more challenging than the use of the other. For the learning algorithms, BNVB mention
that the results tend to be better for the screen inputs. Experiments were
run on a cluster, where each computing node consists of a $6$-core Intel Xeon 
E5-2440, with $2.4$ GHz clock speed, with 64 GBytes of RAM installed.

@@

Last, in the lookahead, \IW(1) and \BFS\ select an action every $5$
frames, while \UCT\ selects an action every frame.  This means that in
order to explore a branch $300$ frames deep, UCT gets to choose $300$
actions, while \IW($1$) and \BFS\ get to choose $60$ actions, both
however using the same $300$ frames from the budget. For this, we
followed the setup of \BRFS\ in BNVB that also selects actions every
$5$ frames, matching the behavior of the emulator that requests an
action also every $5$ frames. Since the lookahead budget is given by a
maximum number of (new) frames, and the time is mostly taken by calls
to the emulator, this may not be the best choice for $\IW(1)$ and
\BFS\ that may therefore not be exploiting all the options afforded by
the budget. Interestingly, when \UCT\ is limited to one action every
$5$ frames, its performance is reduced by up to a $50$\% in games
where it performs very well (\CrazyClimber), and does not appear to
improve in those games where it performs very poorly (\Freeway).

@@

\noindent }

\section*{Acknowledgments} 
We thank the people who created the ALE.  Nir Lipovetzky is partially
funded by the ARC Linkage grant LP11010015.
